Bioinformatics analysis of whole-genome bisulfite-sequencing data

### 2014-12-06 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Number of CpGs per read

Using a tool such as `Trim Galore!` ([http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/](http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/)) or `cutadapt` \cite{Martin:2011va}, the reads can be _trimmed_ to remove these contaminants. Reads might also be trimmed to remove low quality positions, which are common at the 3' end of reads, although this isn't as essential as trimming to remove contaminants.

@@

Read mapping is complicated by the bisulfite-treatment of the DNA. Following bisulfite-treatment, the DNA fragments are now mostly composed of three bases rather than four, which means there are many more sequence mismatches between a read and its true mapping location. Simply using standard read mapping software and allowing for more mismatches would result in many reads mapping to multiple locations in the reference genome. Instead, a field of read mapping software dedicated to bisulfite-sequencing data has developed. Several review articles have tried to summarise and compare the various approaches \cite{Chatterjee:2012ft, Krueger:2012ks, KundeRamamoorthy:2014fj}.

@@

While methylation-aware mappers provide the highest efficiency, these suffer from a bias whereby methylated reads are preferentially mapped over unmethylated reads \cite{Krueger:2012ks}. This biases downstream inference and means that these mappers are generally less popular. I will instead focus on the _in silico_ bisulfite-conversion mappers, called as such because of a key step taken by these in the mapping process.

_In silico_ bisulfite-conversion mappers convert all cytosines to thymines (resp. guanines to adenines) of the forward (resp. reverse) strand from the reference genome. They then take each read and create two _in silico_ bisulfite-converted versions of it[^2_reads]: the CT-read replaces all residual thymines with cytosines and the GA-read replaces all residual guanines with adenines. The CT-read is mapped against the CT-genome and the GA-read is mapped against the GA-genome using a standard mapping tools such as `Bowtie1` \cite{Langmead:2009fv}, `Bowtie2` \cite{Langmead:2012jh} or `bwa`  \cite{Li:2009fi, Li:2010bl}[^non_directional].

@@

_In silico_ bisulfite-conversion mappers avoid the bias inherent in the methylation-aware mappers because all reads, regardless of methylation status, 'look the same' to the mapper. However, they do suffer from a slight loss in mapping efficiency \cite{Krueger:2012ks}.

@@

\begin{table}[h]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
Name      & Reference                                            & Underlying mapping software \\ \midrule
Bismark   & \cite{Krueger:2011eb}                              & Bowtie1 or Bowtie2          \\
bwa-meth  & \cite{Pedersen:2014wt}                             & bwa-mem                     \\
BSMAP     & \cite{Xi:2009bd}                                   & SOAP                        \\
Novoalign & \url{http://www.novocraft.com/products/novoalign/} & Novoalign                   \\ \bottomrule
\end{tabular}
\caption{Four popular bisulfite-sequencing read mappers, selected to highlight the variety of back-ends used by these software.}
\label{tab:bs-seq_mappers}
\end{table}

Each of these aligners can report the output in the standard Sequence Alignment/Map format (`SAM`) or it's binary equivalent, `BAM` \cite{Li:2009ka}. However, there is no agreed upon standard in the `SAM` specification for encoding the data specific to bisulfite-sequencing, which means that each mapper does this in a slightly different way. This makes it difficult to develop downstream analysis tools.

@@

Technically, reads from Illumina sequencing are often too short to resolve the mapping location of these repetitive sequences. Resolving the mapping location of repetitive sequences can be achieved by using other sequencing technologies, such as Pacific Biosciences SMRT technology \cite{Flusberg:2010hra}, which produces longer reads.

@@

Ideally, mapping software assigns the degree of confidence it has that the read is correctly mapped via a mapping quality score (`mapQ`). In theory, reads might be down-weighted in downstream analyses based on the mapping quality score. However, these mapping quality scores are often poorly calibrated, particularly for methylC-seq data, which makes them less useful. `Bismark` \cite{Krueger:2011eb}, a popular bisulfite-sequencing mapping software, only recently introduced mapping quality scores (`v0.12.1` released in April 2014).

The above problems are general challenges of read mapping and are not specific to bisulfite-sequencing data, although the reduced complexity of bisulfite-sequencing reads exacerbates these issues. The difficulty of mapping to repetitive regions of the genome is a particularly frustrating one for  bisulfite-sequencing data. Repetitive sequences, such as LINEs and SINEs, are frequently methylated to prevent transcription and so are of interest to researchers studying DNA methylation. The low mapping efficiencies of these regions means that there is often limited data of these elements from bisulfite-sequencing data (__TODO: CHECK IN MY DATA AND REPORT COVERAGE__).

@@

Generally, there is no way to tell based on sequencing data if a read is truly a PCR duplicate. However, it is relatively easy to identify suspected PCR duplicates (which are almost always inaccurately referred to as "PCR duplicates"[^pcr_dup]). Software to identify PCR duplicates include the `MarkDuplicates` function that is a part of the `Picard` software ([http://broadinstitute.github.io/picard/](http://broadinstitute.github.io/picard/)), the `rmdup` function that is a part of the `SAMtools` software \cite{Li:2009ka} and `SAMBLASTER` \citep{Faust:2014hf}. These software use mapped reads from a `SAM/BAM` file as input. There are subtle differences in how each method calls PCR duplicates. As an example, this is how `Picard`'s `MarkDuplicates` function works:

@@

The loss of the one-to-one relationship between sequencing cycle and read-position cannot be avoided if reads are pre-trimmed because the trimming information is not preserved. \cite{Hansen:2012gr} suggest a separate M-bias plot for each read-length, which will help mitigate the effect of confounding between read-position and sequencing-cycle[^read_trimming]. However, if trimming is performed during the alignment then all the necessary information is retained and the x-axis of M-bias plots would be 'sequencing cycle' rather than 'read-position', thus avoiding the issue entirely.

@@

In practice, the M-bias curves for each sample are visually inspected to look for evidence of M-bias, i.e., read-positions whose methylation levels are 'too far away' from the vast majority of read-positions. A minor problem with this approach is that the user may (unconsciously) use a different cutoff for each sample. For example, it is more difficult to determine an appropriate cutoff when there is a gradual slope in the M-bias plot than when there is a clear jump (__TODO: An example where there last few read-positions are more than `cutoff` from the median but due to a gradual slope vs. an example where a single position is more than `cutoff` from the median due to a spike__). This motivated me to include functionality in my `MethylationTuples` R package ([https://github.com/PeteHaitch/MethylationTuples](https://github.com/PeteHaitch/MethylationTuples))  to perform more systematic processing of M-bias results. These are implemented in the `MBias` class, its associated methods, `trim` and `plot`, and the helper function `readMBias`.

@@

Apart from M-bias, technical biases include sequencing and alignment errors, which can both give rise to spurious or otherwise incorrect methylation calls in obvious ways. Biological biases include cellular heterogeneity and sequence variation at or nearby to methylation loci. I discuss the latter in __TODO: WHERE???__

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
\caption{`Bis-SNP` is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand. For all three loci, the reads mapped to the forward strand contain a thymine. However, it is the base on the reverse strand that reveal the true genotype. When combined with the reference genome it can be inferred whether the sample's genome, which isn't directly observed, has a genetic variant at that location. This is only possible with bisulfite-data generated using the directional protocol. This figure is adapted from \cite{Liu:2012ge}.}
\label{fig:Bis-SNP}
\end{figure}

@@

__TODO: Size of m-tuples histogram__

@@

\begin{table}[h]
\centering
\begin{tabular}{@{}lllll@{}}
\toprule
Software           & Reference                                & Input                                              & Designed for targeted sequencing & Other limitations                                                                \\ \midrule
\texttt{methclone} & \cite{Li:2014ei}                       & Bismark BAM files                                  & No, although example uses eRRBS  & \textbf{TODO: Try installing. Suspect you can't get actual methylation calls.} \\
\texttt{methpat}  & \url{https://github.com/bjpop/methpat} & Output of \texttt{bismark\_methylation\_extractor} & Yes (amplicons)                  &                                                                                  \\
\texttt{DMEAS}     & \cite{He:2013cj}                       & Output of \texttt{bismark\_methylation\_extractor} &                                  & Windows operating system only. Perl code only available as PDF file.             \\ \bottomrule
\end{tabular}
\caption{My caption}
\label{tab:other_m-tuple_methylation_callers}
\end{table}

@@

__TODO: Show example of `methtuple` output__

@@

Figure \ref{fig:methtuple_runtime} shows the distribution of running times and Figure \ref{fig:methtuple_memory} the maximum memory usage across all the samples from the EPISCOPE, Lister and Ziller datasets. For each sample, each chromosome was processed using a single core on one of the shared-use servers in the Bioinformatics Division (see __APPENDIX__ for details of these machines).

>>>>

Using a tool such as `Trim Galore!` ([http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/](http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/)) or `cutadapt` \citep{Martin:2011va}, the reads can be _trimmed_ to remove these contaminants. Reads might also be trimmed to remove low quality positions, which are common at the 3' end of reads, although this isn't as essential as trimming to remove contaminants.

@@

Read mapping is complicated by the bisulfite-treatment of the DNA. Following bisulfite-treatment, the DNA fragments are now mostly composed of three bases rather than four, which means there are many more sequence mismatches between a read and its true mapping location. Simply using standard read mapping software and allowing for more mismatches would result in many reads mapping to multiple locations in the reference genome. Instead, a field of read mapping software dedicated to bisulfite-sequencing data has developed. Several review articles have tried to summarise and compare the various approaches \citep{Chatterjee:2012ft, Krueger:2012ks, KundeRamamoorthy:2014fj}.

@@

While methylation-aware mappers provide the highest efficiency, these suffer from a bias whereby methylated reads are preferentially mapped over unmethylated reads \citep{Krueger:2012ks}. This biases downstream inference and means that these mappers are generally less popular. I will instead focus on the _in silico_ bisulfite-conversion mappers, called as such because of a key step taken by these in the mapping process.

_In silico_ bisulfite-conversion mappers convert all cytosines to thymines (resp. guanines to adenines) of the forward (resp. reverse) strand from the reference genome. They then take each read and create two _in silico_ bisulfite-converted versions of it[^2_reads]: the CT-read replaces all residual thymines with cytosines and the GA-read replaces all residual guanines with adenines. The CT-read is mapped against the CT-genome and the GA-read is mapped against the GA-genome using a standard mapping tools such as `Bowtie1` \citep{Langmead:2009fv}, `Bowtie2` \citep{Langmead:2012jh} or `bwa`  \citep{Li:2009fi, Li:2010bl}[^non_directional].

@@

_In silico_ bisulfite-conversion mappers avoid the bias inherent in the methylation-aware mappers because all reads, regardless of methylation status, 'look the same' to the mapper. However, they do suffer from a slight loss in mapping efficiency \citep{Krueger:2012ks}.

@@

\begin{table}[h]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
Name      & Reference                                            & Underlying mapping software \\ \midrule
Bismark   & \citet{Krueger:2011eb}                              & Bowtie1 or Bowtie2          \\
bwa-meth  & \citet{Pedersen:2014wt}                             & bwa-mem                     \\
BSMAP     & \citet{Xi:2009bd}                                   & SOAP                        \\
Novoalign & \url{http://www.novocraft.com/products/novoalign/} & Novoalign                   \\ \bottomrule
\end{tabular}
\caption{Four popular bisulfite-sequencing read mappers, selected to highlight the variety of back-ends used by these software.}
\label{tab:bs-seq_mappers}
\end{table}

Each of these aligners can report the output in the standard Sequence Alignment/Map format (`SAM`) or it's binary equivalent, `BAM` \citep{Li:2009ka}. However, there is no agreed upon standard in the `SAM` specification for encoding the data specific to bisulfite-sequencing, which means that each mapper does this in a slightly different way. This makes it difficult to develop downstream analysis tools.

@@

Technically, reads from Illumina sequencing are often too short to resolve the mapping location of these repetitive sequences. Resolving the mapping location of repetitive sequences can be achieved by using other sequencing technologies, such as Pacific Biosciences SMRT technology \citep{Flusberg:2010hra}, which produces longer reads.

@@

Ideally, mapping software assigns the degree of confidence it has that the read is correctly mapped via a mapping quality score (`mapQ`). In theory, reads might be down-weighted in downstream analyses based on the mapping quality score. However, these mapping quality scores are often poorly calibrated, particularly for methylC-seq data, which makes them less useful. `Bismark` \citep{Krueger:2011eb}, a popular bisulfite-sequencing mapping software, only recently introduced mapping quality scores (`v0.12.1` released in April 2014).

The above problems are general challenges of read mapping and are not specific to bisulfite-sequencing data, although the reduced complexity of bisulfite-sequencing reads exacerbates these issues. The difficulty of mapping to repetitive regions of the genome is a particularly frustrating one for  bisulfite-sequencing data. Repetitive sequences, such as LINEs and SINEs, are typically methylated in order to prevent transcription and are often of interest to researchers studying DNA methylation. The low mapping efficiencies of these regions means that there is often limited or less reliable data for these elements from bisulfite-sequencing data.

@@

Generally, there is no way to tell based on sequencing data if a read is truly a PCR duplicate. However, it is relatively easy to identify suspected PCR duplicates (which are almost always inaccurately referred to as "PCR duplicates"[^pcr_dup]). Software to identify PCR duplicates include the `MarkDuplicates` function that is a part of the `Picard` software ([http://broadinstitute.github.io/picard/](http://broadinstitute.github.io/picard/)), the `rmdup` function that is a part of the `SAMtools` software \citep{Li:2009ka} and `SAMBLASTER` \citep{Faust:2014hf}. These software use mapped reads from a `SAM/BAM` file as input. There are subtle differences in how each method calls PCR duplicates. As an example, this is how `Picard`'s `MarkDuplicates` function works:

@@

The loss of the one-to-one relationship between sequencing cycle and read-position cannot be avoided if reads are pre-trimmed because the trimming information is not preserved. \citet{Hansen:2012gr} suggest a separate M-bias plot for each read-length, which will help mitigate the effect of confounding between read-position and sequencing-cycle[^read_trimming]. However, if trimming is performed during the alignment then all the necessary information is retained and the x-axis of M-bias plots would be 'sequencing cycle' rather than 'read-position', thus avoiding the issue entirely.

@@

In practice, the M-bias curves for each sample are visually inspected to look for evidence of M-bias, i.e., read-positions whose methylation levels are 'too far away' from the vast majority of read-positions. It is rather a subjective decision to make. Two problems that I found were in maintaining consistency across samples and in determining where to draw the line when there is no dramatic 'spike' in the M-bias. For example, in Figure \ref{fig:E18VA_M-bias} it is clear that there are problems at the start and end of both read 1 and read 2, as well as a big problem at read-position 101 in read 2. What is less clear is where to draw the line on the gradual decay downwards toward the end of read 1 and towards the start of read 2. This motivated me to write a few simple functions to do perform more systematic processing of M-bias results. These are included in the `MethylationTuples` R package, described in Section \ref{sec:MethylationTuples}, with the `MBias` class, its associated methods, `trim` and `plot`, and the helper function `readMBias`.

@@

Apart from M-bias, technical biases include sequencing and alignment errors, which can both give rise to spurious or otherwise incorrect methylation calls in obvious ways. Biological biases include cellular heterogeneity and sequence variation at or nearby to methylation loci.

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
\caption{`Bis-SNP` is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand. For all three loci, the reads mapped to the forward strand contain a thymine. However, it is the base on the reverse strand that reveal the true genotype. When combined with the reference genome it can be inferred whether the sample's genome, which isn't directly observed, has a genetic variant at that location. This is only possible with bisulfite-data generated using the directional protocol. This figure is adapted from \citet{Liu:2012ge}.}
\label{fig:Bis-SNP}
\end{figure}

@@

The number of CpGs per read for the Lister data is shown in Figure \ref{fig:Lister_cpgs_per_read}. Longer reads, and paired-end reads, contain more methylation loci and so are more informative for analyses using m-tuples. This can be seen by comparing, for example, the _ADS_ and _HSF1_ samples. Samples sequenced more deeply will have more reads per m-tuple, although this can't be seen in these plots since they are normalised by sequencing depth.

```{r cpgs_per_read, fig.lp = "fig:Lister_cpgs_per_read", dev = "pdf", eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "Number of CpGs per read for the Lister data."}
library(dplyr)
library(ggplot2)
library(scales)
thesis_theme <- theme_classic(base_size = 20)
thesis_theme <- theme_bw(base_size = 20)
dataset <- "Lister"
hist <- lapply(paste0("../data/observable_m-tuples/", dataset, "/",
                      list.files(paste0("../data/observable_m-tuples/",
                                        dataset), "\\.CG\\_per\\_read\\.hist")),
               read.table, header = TRUE)
sn <- gsub("\\.CG\\_per\\_read\\.hist", "",
           list.files(paste0("../data/observable_m-tuples/",
                             dataset), "\\.CG\\_per\\_read\\.hist"))
sn <- gsub("DM\\_", "", sn)
hist_df <- bind_rows(hist) %>%
  mutate(Sample = rep(sn, sapply(hist, nrow))) %>%
  group_by(Sample) %>%
  mutate(Proportion = count / sum(as.numeric(count)))
hist_df %>%
  ggplot(aes(x = n, y = Proportion * 100)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Sample, ncol = 5, nrow = 4) +
  coord_cartesian(xlim = c(-1, 8.5)) +
  scale_y_continuous(label = comma) +
  xlab("Number of CpGs per read\n(Axis truncated at 8)") +
  ylab("Percentage of total reads") +
  ggtitle(paste0(dataset, ": Number of CpGs per read")) +
  thesis_theme
```

@@

\begin{table}[h]
\centering
\begin{tabular}{@{}lllll@{}}
\toprule
Software           & Reference                                & Input                                              & Designed for targeted sequencing & Other limitations                                                                \\ \midrule
\texttt{methclone} & \citet{Li:2014ei}                       & Bismark BAM files                                  & No, although example uses eRRBS  & \textbf{TODO: Try installing. Suspect you can't get actual methylation calls.} \\
\texttt{methpat}  & \url{https://github.com/bjpop/methpat} & Output of \texttt{bismark\_methylation\_extractor} & Yes (amplicons)                  &                                                                                  \\
\texttt{DMEAS}     & \citet{He:2013cj}                       & Output of \texttt{bismark\_methylation\_extractor} &                                  & Windows operating system only. Perl code only available as PDF file.             \\ \bottomrule
\end{tabular}
\caption{My caption}
\label{tab:other_m-tuple_methylation_callers}
\end{table}

@@

@@

Figure \ref{fig:methtuple_runtime} shows the distribution of running times and Figure \ref{fig:methtuple_memory} the maximum memory usage across all the samples from the EPISCOPE, Lister and Ziller datasets. For each sample, each chromosome was processed using a single core on one of the shared-use servers in the Bioinformatics Division (see Table \ref{tab:unices} in the Appendix for details of these machines).
